$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-05-01 paper/paper.tex
Go through the changes, until 4.1

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Syntax}
\label{syntax}
We let $x, y,\ldots$ range over \emph{channel names} or \emph{variables} and
$P, Q,\ldots$ over processes.
The syntax of the \picalc{} \cite{Sangio01} is given by the following grammar.
\todo{maybe put both syntaxes in two columns to save space}
\begin{equation*}
  \begin{split}
    P,Q ::=      & \; \PO       &\text{(terminated process)}  \\
                |& \; (\new{}x)  P   & \text{(restriction)} \\
                |& \; \comp{P}{Q}  & \text{(parallel composition)} \\
                |& \; \recvp{x}{y}. \; P  & \text{(input)} \\
                |& \; \sendp{x}{y}. \; P   &\text{(output)}
  \end{split}
\end{equation*}
Process $\PO$ denotes the terminated process, where no operations can further occur.
Process $(\new{}x)P$ is the restriction process where channel $x$ is created new and is bound with scope $P$.
Process $\comp{P}{Q}$ is the parallel composition of processes $P$ and $Q$.
Processes $\recvp{x}{y}.P$ and $\sendp{x}{y}.P$ denote respectively the input and output processes of a variable/channel $y$ over a channel $x$, with continuation $P$.

Scope restriction $(\new{}x)P$ and input $\recvp{x}{y}.P$ both introduce bound names in $P$.
In order to mechanise the \picalc{} syntax in Agda, we need a way to deal with bound names in these continuation processes.
Channel names are cumbersome to mechanise: inserting a new variable into a context means proving that the name of such variable is different to all other variable names in context.
Instead we adopt de Bruijn indices \cite{} where a natural number $n$ (aka \emph{index}) is used to refer to the variable introduced $n$ binders ago --- hence binders no longer introduce names.
\begin{example}
\hfill{}\\
The \picalc{} term \hfill{} $(\new{}x ) (\comp {\recvp{x}{y} . \; \sendp{y}{\texttt{"Hello world!"}} . \; \PO} {(\new{} y) (\sendp{x}{y} . \; \recvp{y}{z} . \; \PO)})$ \\ is represented as \hfill{} $\new{} (\comp {\recv{0} . \; \send{0}{\texttt{"Hello world!"}} . \; \PO} {\new{} (\send{1}{0} . \; \recv{0} . \; \PO)})$.
\end{example}
That is, terms at different \emph{depths} use different indices to refer to the same binding.

@@

Every time we go under a binder, the number of binders a variable might refer to increments by one.
To propagate this information, we index processes according to their \emph{depth}: for all naturals $n$, a process of type $\Process_n$ contains variables that can refer to $n$ distinct elements.
\autoref{fig:syntax} shows the syntax of \picalc{} in Agda. We increase the \emph{depth} counter by 1 in channel restriction and input processes.

@@

Using type-level de Bruijn indices makes our syntax well-scoped by construction --- the semantics of our language can now be defined on the totality of the syntax.
In order to make the de Bruijn indices representation more human-readable, we can annotate these indices with the original channel names.
Machines can then manipulate references mechanically whilst using channel names to present \picalc{} processes back to humans.

@@

Structural congruence $\eqeq$ is fully defined in \autoref{app:struct}.
\autoref{fig:struct-cong-base} lists the base cases.

@@

  \inferrule
  {uQ : \Unused_0 \, Q}
  {\constr{scope-ext} : \new (\comp{P}{Q}) \eqeq \comp{(\new P)}{lower_0 \; \; Q \; uQ}}

@@

The first three rules on $\constr{comp}$ state respectively, associativity, symmetry, and $\PO$ as the neutral element for parallel composition.
The last three rules on $\constr{scope}$ state respectively, garbage collection, scope extrusion and commutativity of restrictions.

In $\constr{scope-ext}$ the type $\Unused_i \; Q$ is an inductive proof that witnesses that the variable index $i$ does not appear neither in the inputs nor in the outputs of $Q$.
The function $lower_i \; Q \; uQ$ traverses $Q$ decrementing every index bigger than $i$.
In $\constr{scope-comm}$ function $swap_i \; P$ traverses $P$ (of type $\Process_{\suc \suc n}$) and swaps variable references $i$ and $\suc i$.
In all the above, $i$ is incremented every time we go under a binder.

@@

The operational semantics of the \picalc{} is defined via a reduction relation $\reduce{}$ in \autoref{fig:reduction}.
In rule $\constr{comm}$, processes in parallel reduce when they communicate over the same index ${i}$.
The input process $\recv{i}P$ substitutes in $P$ references to the most immediate variable placeholder with references to the variable sent by the output process $\send{i}{j}{Q}$.
The continuation $\subst{P}{j}{0}$ is then \emph{lowered} --- all variable references are decreased by one.
Reduction is closed under parallel composition (rule $\constr{par}$), restriction (rule $\constr{res}$) and structural congruence (rule $\constr{struct}$) 
--- notably, not under input or output, otherwise the sequencing of actions would not be preserved \cite{Sangio01}.
%In  \autoref{subject-reduction} we prove that if $P$ reduces to $Q$ and $P$ is well-typed, so is $Q$.

\todo{this paragraph below is not clear}
Notice that reduction is effectful: it consumes the channel over which communication occurs.
If the channel is $\constr{external}$ to $P$ (it resides in its typing context) then the context in which $Q$ is typed must change.
To keep track of this information, we lift the index of the channel to the type level.
We use a $dec$ function to decrement the index of the channel every time we come out of a binder.
This function saturates at $\constr{internal}$ --- meaning communication happens internally to the process.

@@

A typing context $\Gamma$ is a partial function from channel names to types.
A typing judgment is of the form $\Gamma \vdash P$, reading ``process $P$ is well typed under the typing context $\Gamma$''.
A resource-aware typing system for the \picalc{} must take resource usages into account, and needs an algebra on these --- a way of \emph{splitting} resources on parallel composition, of \emph{ticking off} resources on input and output.
We characterise such an algebra in  \autoref{multiplicities}.
We define independent typing contexts and usage contexts in  \autoref{contexts}.
We provide a typing system with leftovers for a resource-aware \picalc{} in  \autoref{leftover-typing}.

@@

\todo{note I've changed the order of presentation from linear to shared, check if it breaks anything!}
Linear, gradual and shared types are all defined as an instance of the laws in \autoref{fig:multiplicities}.
We illustrate these types in typing derivations in \autoref{example-derivations}.

@@

\subsection{Typing Contexts}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\label{contexts}
\todo{is usage annotation the same as multiplicities? if yes, just use the former as too many notions}
In a resource-aware \picalc{}, usage annotations are transformed as a result of input and output; types however are preserved.
In our case, usage annotations do not depend on types, and keeping types and usages in separate contexts allows for a more modular approach in our proofs.

@@

\begin{note}
  We lift the monoidal operation on multiplicity pairs $\opsquared{x}{y}{z}$ to a monoidal operation $\opctx{\Gamma}{\Delta}{\Xi}$ on contexts of usages with the same underlying context of indices.
  All the properties of the partial monoid are trivially lifted.
\end{note}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Syntax}
\label{syntax}
Let $x, y,\ldots$ range over \emph{variables} and $P, Q,\ldots$ over processes.
The syntax of the \picalc{} \cite{Sangio01} is given by the following grammar.
\begin{equation*}
  \begin{split}
    P,Q ::=      & \; \PO       &\text{(terminated process)}  \\
                |& \; (\new{}x)  P   & \text{(restriction)} \\
                |& \; \comp{P}{Q}  & \text{(parallel composition)} \\
                |& \; \recvp{x}{y}. \; P  & \text{(input)} \\
                |& \; \sendp{x}{y}. \; P   &\text{(output)}
  \end{split}
\end{equation*}
Process $\PO$ denotes the terminated process, where no operations can further occur.
Process $(\new{}x)P$ creates a new channel $x$ bound with scope $P$.
Process $\comp{P}{Q}$ is the parallel composition of processes $P$ and $Q$.
Processes $\recvp{x}{y}.P$ and $\sendp{x}{y}.P$ denote respectively the input and output processes of a variable $y$ over a channel $x$, with continuation $P$.

Scope restriction $(\new{}x)P$ and input $\recvp{x}{y}.P$ both introduce bound names in $P$.
In order to mechanise the \picalc{} syntax in Agda, we need to deal with bound names in continuation processes.
Names are cumbersome to mechanise: inserting a new variable into a context means proving that its name differs from all other variable names in context.
Instead, we use de Bruijn indices \cite{}, where a natural number $n$ (aka \emph{index}) is used to refer to the variable introduced $n$ binders ago --- hence binders no longer introduce names.
\begin{example}
\hfill{}\\
The \picalc{} term \hfill{} $(\new{}x ) (\comp {\recvp{x}{y} . \; \sendp{y}{\texttt{"Hello world!"}} . \; \PO} {(\new{} y) (\sendp{x}{y} . \; \recvp{y}{z} . \; \PO)})$ \\ is represented as \hfill{} $\new{} (\comp {\recv{0} . \; \send{0}{\texttt{"Hello world!"}} . \; \PO} {\new{} (\send{1}{0} . \; \recv{0} . \; \PO)})$.
\end{example}
That is, terms at different \emph{depths} use different indices to refer to the same binding.

@@

Every time we go under a binder, the number of binders a variable might refer to increments by one.
To propagate this information, we index processes according to their \emph{depth}: for all naturals $n$, a process of type $\Process_n$ contains variables that can refer to $n$ distinct elements.
We increase the \emph{depth} counter by 1 in channel restriction and input processes.
\autoref{fig:syntax} showcases a \picalc{} syntax that is well-scoped by construction thanks to type-level de Bruijn indices.

@@

The semantics of our language can now be defined on the totality of the syntax.

@@

We define the base cases of a structural congruence relation \textsc{StructCong} in \autoref{fig:struct-cong-base}.
Its congruent equivalence closure \textsc{Equals} is provided in \autoref{app:struct}.

@@

  \inferrule
  {uQ : \Unused_0 \; Q}
  {\constr{scope-ext} : \new (\comp{P}{Q}) \eqeq \comp{(\new P)}{lower_0 \; \; Q \; uQ}}

@@

The first three rules ($\constr{comp}-$*) state associativity, symmetry, and $\PO$ as the neutral element of parallel composition, respectively.
The last three ($\constr{scope}-$*) state, garbage collection, scope extrusion and commutativity of restrictions, respectively.

In $\constr{scope-ext}$ the type $\Unused_i \; Q$ is an inductive proof asserting that the variable index $i$ does not appear neither in the inputs nor in the outputs of $Q$.
The function $lower_i \; Q \; uQ$ traverses $Q$ decrementing every index bigger than $i$.
In $\constr{scope-comm}$ the function $swap_i \; P$ traverses $P$ (of type $\Process_{\suc \suc n}$) and swaps variable references $i$ and $\suc i$.
In all the above, $i$ is incremented every time we go under a binder.

@@

The operational semantics of the \picalc{} is defined as a reduction relation \textsc{Reduces} indexed by the channel $i$ on which communication occurs (\autoref{fig:reduction}).
In rule $\constr{comm}$, parallel processes reduce when they communicate over a common channel ${i}$.
As a result of that communication, the continuation of the input process $P$ gets all the references to its most immediate variable substituted with references to $\suc j$, the variable sent by the output process $\send{i}{j}{Q}$.
After substitution, $\subst{P}{j}{0}$ is \emph{lowered} --- all variable references are decreased by one.
Reduction is closed under parallel composition (rule $\constr{par}$), restriction (rule $\constr{res}$) and structural congruence (rule $\constr{struct}$) 
--- notably, not under input or output, otherwise the sequencing of actions would not be preserved \cite{Sangio01}.

We keep track of the channel over which communication occurs so that we can state subject reduction.
We use a $dec$ function to decrement the index of this channel every time we come out of a binder.
This function saturates at $\constr{internal}$ --- meaning communication happens internally to the process.

@@

A resource-aware typing system for the \picalc{} must take resource usages into account, and needs an algebra on these --- a way of \emph{splitting} resources on parallel composition, of \emph{ticking off} resources on input and output.
We characterise such an algebra in  \autoref{multiplicities}.
We define independent typing contexts and usage contexts in  \autoref{contexts}.
We provide a typing system with leftovers for a resource-aware \picalc{} in  \autoref{leftover-typing}.

@@

Linear, gradual and shared types are all defined as an instance of the laws in \autoref{fig:multiplicities}.
We illustrate these types in typing derivations in \autoref{example-derivations}.

@@

\subsection{Typing Contexts}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\label{contexts}
In a resource-aware \picalc{}, usage annotations are transformed as a result of input and output; types however are preserved.
In our case, usage annotations do not depend on types, and keeping types and usages in separate contexts allows for a more modular approach in our proofs.

@@

\begin{note}
  We lift the monoidal operation on usage annotations $\opsquared{x}{y}{z}$ to a monoidal operation $\opctx{\Gamma}{\Delta}{\Xi}$ on usage contexts with a common underlying context of indices.
  All the properties of the partial monoid are trivially lifted.
\end{note}
